### 2018-07-03 paper/1-intro.tex
Final tweaks

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.  Moreover,
For years \Make{} dominated, but more recently
Recently, the challenges of scale have driven large software firms
like Microsoft, Facebook, and Google to develop their own build
systems, exploring new points in the design space.
These complex build systems use subtle algorithms, but they
are often hidden away, and not the object of study.

>>>>

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
% Rarely do people ask questions like ``What does it mean for my build
% system to be correct?'' or ``What are the trade-offs between different
% approaches?''.  Moreover,
For years \Make dominated, but more recently the challenges of scale have driven
large software firms like Microsoft, Facebook, and Google to develop their own
build systems, exploring new points in the design space. These complex build
systems use subtle algorithms, but they are often hidden away, and not the
object of study.

### 2018-07-03 paper/2-background.tex
Final tweaks

Item (iii) in the above list highlights another distinguishing feature of \Excel: it
is \emph{self-tracking}. Most build systems only track changes of inputs and
intermediate results, but \Excel also tracks changes in the tasks themselves: if
a formula is modified, \Excel will recompute it and propagate the changes.
Self-tracking is uncommon in software build systems, where one often needs to
manually initiate a full rebuild even if just a single build task has changed.
We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

>>>>

Item (iii) in the above list highlights another distinguishing feature of \Excel: it
is \emph{self-tracking}. Most build systems only track changes of inputs and
intermediate results, but \Excel also tracks changes in the tasks themselves: if
a formula is modified, \Excel will recompute it and propagate the changes.
Self-tracking is uncommon in software build systems, where one often needs to
manually initiate a full rebuild even if just a single task has changed.
We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

### 2018-07-03 paper/4-build.tex
Final tweaks

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can try to incrementalise the
topological sort by storing the topological order between build runs,
using it as a starting point in the next run, 
fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT}
stores only hashes, of type \hs{Hash}~\hs{v}, of those values.  Since the
verfying trace persists from one build to the next -- it constitutes the build system's ``memory''
-- it is helpful for it to be of modest size.
After successfully building a key, we call \hs{recordVT} to add a record
to the current \hs{VT}, passing the key, the hash of
its value, and the list of hashes and dependencies.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well
if we skip any number dependency levels (say $n$ levels).
The input-only approach is the special
case of $n = \infty$, and constructive trances are the special case of 
$n = 1$.

@@

With the information in this section we can build Table~\ref{tab-build-systems}, which
tabulates combinations of the schedling strategy and the out-of-date keys strategy,
providing 12~possible build systems, 8~of which are actually inhabited by existing build systems
(we discuss all these systems in \S\ref{sec-related-build}). Of the
remaining 4~spots, all result in workable build systems. The most interesting
unfilled spot in the table is suspending constructive traces, which would provide
many benefits, and which we title \Cloud \Shake
% (as we plan on extending \Shake to occupy that spot)
and further explore further in~\S\ref{sec-implementation-cloud}.

>>>>

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can incrementalise the topological
sort by storing the topological order between build runs, using it as a starting
point in the next run, fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} stores
only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
trace persists from one build to the next -- it constitutes the build system's
``memory'' -- it is helpful for it to be of modest size. After successfully
building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
passing the key, the hash of its value, and the list of hashes and dependencies.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number $n$ of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

@@

With the information in this section we can build Table~\ref{tab-build-systems},
which tabulates combinations of the scheduling strategy and the out-of-date keys
strategy, providing 12~possible build systems, 8~of which are actually inhabited
by existing build systems (we discuss all these systems in
\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table is suspending
constructive traces, which would provide many benefits, and which we title
\Cloud \Shake and further explore further in~\S\ref{sec-implementation-cloud}.
% (as we plan on extending \Shake to occupy that spot)

### 2018-07-03 paper/5-implementations.tex
Final tweaks

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types (Figure~\ref{fig-types})
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The \hs{suspending} scheduler is based on a recursive function \hs{build} that
builds a given \hs{key}, making sure not to duplicate work when called on the
same \hs{key} again in future. To achieve that, it keeps track of keys that have
already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}, running the
computation in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.

>>>>

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types (Fig.~\ref{fig-types}):

\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}

\noindent
A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The \hs{suspending} scheduler uses a recursive function \hs{build} that builds
a given \hs{key}, making sure not to duplicate work when called on the same
\hs{key} again in future. To achieve that, it keeps track of keys that have
already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}.
% , running the computation in the
% \hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.

### 2018-07-03 paper/7-related.tex
Final tweaks

To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed.

@@

\begin{minted}[xleftmargin=10pt]{haskell}
data Key = A Integer | B Integer | D Integer Integer deriving Ord
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
editDistance :: Task Monad Key Integer
editDistance _     (D i 0) = Just $ pure i
editDistance _     (D 0 j) = Just $ pure j
editDistance fetch (D i j) = Just $ do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj
        then fetch (D (i - 1) (j - 1))
        else do
            insert  <- fetch (D  i      (j - 1))
            delete  <- fetch (D (i - 1)  j     )
            replace <- fetch (D (i - 1) (j - 1))
            return (1 + minimum [insert, delete, replace])
editDistance _ _ = Nothing
\end{minted}

>>>>

To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The key difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from output to action,
\Fabricate supplies a list of statements, in an order, without declaring what each line produces.
There is no need to schedule the statements, and not enough information to do so.

@@

\begin{minted}[xleftmargin=10pt]{haskell}
data Key = A Integer | B Integer | D Integer Integer deriving Ord
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
editDistance :: Tasks Monad Key Integer
editDistance (D i 0) = Just $ Task $ const $ pure i
editDistance (D 0 j) = Just $ Task $ const $ pure j
editDistance (D i j) = Just $ Task $ \fetch -> do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj then fetch (D (i - 1) (j - 1))
                else do insert  <- fetch (D  i      (j - 1))
                        delete  <- fetch (D (i - 1)  j     )
                        replace <- fetch (D (i - 1) (j - 1))
                        return (1 + minimum [insert, delete, replace])
editDistance _ = Nothing
\end{minted}
